---
title: "Hanihara Dental Project"
author: "Elaine Y. Chu and Tatiana VM"
date: "December 3, 2019"
output: 
  html_document:
     theme: cerulean
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, fig.path='Figures/', fig.ext='jpg', fig.width=8, fig.height=8, dpi=180)

library(readxl)
library(tidyverse)
library(missForest)
library(VIM)
library(imputeMissings)
library(irr)
library(binda)
library(ggdendro)
library(gridExtra)
library(dendextend)

setwd("C:/Users/elain/Box Sync/Projects/w_Tatiana/ImputingScores/Pub")
set.seed(123)
```

## Hanihara Dataset

Here, we chose 4 distinct groups for our analysis: Euro, Aust, Mel, and NAm. 

```{r Data Munge}
og <- read.csv("Hanihara_cleaned_project.csv")

## Choose Max Expression
complete <- og
complete$SHOV_UI2<-pmax(complete$SHOV_UI2_L,complete$SHOV_UI2_R)
complete$MIT_UP1<-pmax(complete$MIT_UP1_L,complete$MIT_UP1_R)
complete$MIT_UP2<-pmax(complete$MIT_UP2_L,complete$MIT_UP2_R)
complete$X6CUP_LM1 <-pmax(complete$X6CUP_LM1_L,complete$X6CUP_LM1_R)
complete$X7CUP_LM1<-pmax(complete$X7CUP_LM1_L,complete$X7CUP_LM1_R)
complete$DTC_LM1<-pmax(complete$DTC_LM1_L,complete$DTC_LM1_R)
complete$CNO_LM2<-pmax(complete$CNO_LM2_L,complete$CNO_LM2_R)
complete$X6CUP_LM2<-pmax(complete$X6CUP_LM2_L,complete$X6CUP_LM2_R)

complete <- complete[,c(2:3,24:31)]  # select only the max expressions
complete[complete==-9] <- NA  # replace -9 with NA

complete <- na.omit(complete)  # remove NAs
complete$Pop <- as.factor(complete$Pop)
complete$Sex <- as.factor(complete$Sex)

summary(complete)

write.csv(complete,"Hanihara_complete_cases.csv")

## Testing Imputation Methods
base <- complete[,c(3:10)]
```

First, we run MMD on the complete dataset, with no missing data.

```{r Complete Dataset}
### Dichotomize Data
threshold <- c(1,1,1,1,1,1,5,5)  # set threshold values for each trait

d_base <- as.data.frame(dichotomize(base, threshold))
pops <- complete$Pop
vars <- colnames(complete[,c(3:10)])
d_base <- cbind(pops,d_base)

## Separate pop groups
Euro_base <- filter(d_base, pops=='Euro')
Aust_base <- filter(d_base, pops=='Australian')
Mel_base <- filter(d_base, pops=='Melanesia')
NAm_base <- filter(d_base, pops=='NA_NatAm')

## Calculate Frequencies
freq_Euro_base <- colMeans(Euro_base[,c(2:9)])
freq_Aust_base <- colMeans(Aust_base[,c(2:9)])
freq_Mel_base <- colMeans(Mel_base[,c(2:9)])
freq_NAm_base <- colMeans(NAm_base[,c(2:9)])

freq_base <- rbind(freq_Euro_base,freq_Aust_base,freq_Mel_base,freq_NAm_base)
rownames(freq_base) <- c("Euro","Aust","Mel","NAm")
write.csv(freq_base, "Freq_Base.csv")

## Calculate Sample sizes
samp_Euro_base <- rep(nrow(Euro_base),ncol(Euro_base)-1)
samp_Aust_base <- rep(nrow(Aust_base),ncol(Aust_base)-1)
samp_Mel_base <- rep(nrow(Mel_base),ncol(Mel_base)-1)
samp_NAm_base <- rep(nrow(NAm_base),ncol(NAm_base)-1)

samp_base <- rbind(samp_Euro_base,samp_Aust_base,samp_Mel_base,samp_NAm_base)
colnames(samp_base) <- vars
rownames(samp_base) <- c("Euro","Aust","Mel","NAm")

write.csv(samp_base, "Samp_Base.csv")

## MMD For Base Data
MMD_base <- read.csv("MMD_Base.csv")
dend_base <- hclust(as.dist(MMD_base[,c(2:5)]), method='ward.D2')     
p1 <- ggdendrogram(dend_base) + labs(title="Complete Dataset")
p1
```

Next, we simulate missing data (50%) to mimic a realistic bioarch situation, and do an MMD on that.

```{r Simulated Dataset}
sim <- prodNA(base, noNA=.5)

### Dichotomize Data
d_sim <- as.data.frame(dichotomize(sim, threshold))
d_sim <- cbind(pops,d_sim)

## Separate pop groups
Euro_sim <- filter(d_sim, pops=='Euro')
Aust_sim <- filter(d_sim, pops=='Australian')
Mel_sim <- filter(d_sim, pops=='Melanesia')
NAm_sim <- filter(d_sim, pops=='NA_NatAm')

## Calculate Frequencies
freq_Euro_sim <- colMeans(Euro_sim[,c(2:9)], na.rm=TRUE)
freq_Aust_sim <- colMeans(Aust_sim[,c(2:9)], na.rm=TRUE)
freq_Mel_sim <- colMeans(Mel_sim[,c(2:9)], na.rm=TRUE)
freq_NAm_sim <- colMeans(NAm_sim[,c(2:9)], na.rm=TRUE)

freq_sim <- rbind(freq_Euro_sim,freq_Aust_sim,freq_Mel_sim,freq_NAm_sim)
rownames(freq_sim) <- c("Euro","Aust","Mel","NAm")
write.csv(freq_sim, "Freq_sim.csv")

## Calculate Sample sizes
samp_Euro_sim <- colSums(!is.na(Euro_sim[,c(2:9)]))
samp_Aust_sim <-  colSums(!is.na(Aust_sim[,c(2:9)]))
samp_Mel_sim <-  colSums(!is.na(Mel_sim[,c(2:9)]))
samp_NAm_sim <-  colSums(!is.na(NAm_sim[,c(2:9)]))

samp_sim <- rbind(samp_Euro_sim,samp_Aust_sim,samp_Mel_sim,samp_NAm_sim)
colnames(samp_sim) <- vars
rownames(samp_sim) <- c("Euro","Aust","Mel","NAm")

write.csv(samp_sim, "Samp_sim.csv")

## MMD For sim Data
MMD_sim <- read.csv("MMD_sim.csv")
dend_sim <- hclust(as.dist(MMD_sim[,c(2:5)]), method='ward.D2')     
p2 <- ggdendrogram(dend_sim) + labs(title="Simulated Missing Data") 
p2
```

Now, we test on an imputed dataset, using KNN. And then we run MMD.

```{r Imputed Dataset}
KNN_all <- kNN(sim,imp_var=F)

### Dichotomize Data
d_KNN_all <- as.data.frame(dichotomize(KNN_all, threshold))
d_KNN_all <- cbind(pops,d_KNN_all)

## Separate pop groups
Euro_KNN_all <- filter(d_KNN_all, pops=='Euro')
Aust_KNN_all <- filter(d_KNN_all, pops=='Australian')
Mel_KNN_all <- filter(d_KNN_all, pops=='Melanesia')
NAm_KNN_all <- filter(d_KNN_all, pops=='NA_NatAm')

## Calculate Frequencies
freq_Euro_KNN_all <- colMeans(Euro_KNN_all[,c(2:9)], na.rm=TRUE)
freq_Aust_KNN_all <- colMeans(Aust_KNN_all[,c(2:9)], na.rm=TRUE)
freq_Mel_KNN_all <- colMeans(Mel_KNN_all[,c(2:9)], na.rm=TRUE)
freq_NAm_KNN_all <- colMeans(NAm_KNN_all[,c(2:9)], na.rm=TRUE)

freq_KNN_all <- rbind(freq_Euro_KNN_all,freq_Aust_KNN_all,freq_Mel_KNN_all,freq_NAm_KNN_all)
rownames(freq_KNN_all) <- c("Euro","Aust","Mel","NAm")
write.csv(freq_KNN_all, "Freq_KNN_all.csv")

## Calculate Sample sizes
samp_Euro_KNN_all <- colSums(!is.na(Euro_KNN_all[,c(2:9)]))
samp_Aust_KNN_all <-  colSums(!is.na(Aust_KNN_all[,c(2:9)]))
samp_Mel_KNN_all <-  colSums(!is.na(Mel_KNN_all[,c(2:9)]))
samp_NAm_KNN_all <-  colSums(!is.na(NAm_KNN_all[,c(2:9)]))

samp_KNN_all <- rbind(samp_Euro_KNN_all,samp_Aust_KNN_all,samp_Mel_KNN_all,samp_NAm_KNN_all)
colnames(samp_KNN_all) <- vars
rownames(samp_KNN_all) <- c("Euro","Aust","Mel","NAm")

write.csv(samp_KNN_all, "Samp_KNN_all.csv")

## MMD For KNN_all Data
MMD_KNN_all <- read.csv("MMD_KNN_all.csv")
dend_KNN_all <- hclust(as.dist(MMD_KNN_all[,c(2:5)]), method='ward.D2')     
p3 <- ggdendrogram(dend_KNN_all) + labs(title="KNN Together") 
p3
```

Yay! Now, we should impute the data by each population and see what it looks like.

```{r Imputed Separately Dataset}
## Separate Datasets
sim_labeled <- cbind(pops, sim)
Euro_sep <- filter(sim_labeled, pops=="Euro")
Aust_sep <- filter(sim_labeled, pops=="Australian")
Mel_sep <- filter(sim_labeled, pops=='Melanesia')
NAm_sep <- filter(sim_labeled, pops=='NA_NatAm')

## Impute Data
Euro_sim <- kNN(Euro_sep,imp_var=F)
Aust_sim <- kNN(Aust_sep,imp_var=F)
Mel_sim <- kNN(Mel_sep,imp_var=F)
NAm_sim <- kNN(NAm_sep,imp_var=F)

### Dichotomize Data
d_Euro_sim <- as.data.frame(dichotomize(Euro_sim[,c(2:9)], threshold))
geo_tag <- rep("Euro",nrow(d_Euro_sim))
d_Euro_sim <- cbind(geo_tag,d_Euro_sim)

d_Aust_sim <- as.data.frame(dichotomize(Aust_sim[,c(2:9)], threshold))
geo_tag <- rep("Aust",nrow(d_Aust_sim))
d_Aust_sim <- cbind(geo_tag,d_Aust_sim)

d_Mel_sim <- as.data.frame(dichotomize(Mel_sim[,c(2:9)], threshold))
geo_tag <- rep("Mel",nrow(d_Mel_sim))
d_Mel_sim <- cbind(geo_tag,d_Mel_sim)

d_NAm_sim <- as.data.frame(dichotomize(NAm_sim[,c(2:9)], threshold))
geo_tag <- rep("NAm",nrow(d_NAm_sim))
d_NAm_sim <- cbind(geo_tag,d_NAm_sim)

## Calculate Frequencies
freq_Euro_sim <- colMeans(d_Euro_sim[,c(2:9)], na.rm=TRUE)
freq_Aust_sim <- colMeans(d_Aust_sim[,c(2:9)], na.rm=TRUE)
freq_Mel_sim <- colMeans(d_Mel_sim[,c(2:9)], na.rm=TRUE)
freq_NAm_sim <- colMeans(d_NAm_sim[,c(2:9)], na.rm=TRUE)

freq_KNN_sep <- rbind(freq_Euro_sim,freq_Aust_sim,freq_Mel_sim,freq_NAm_sim)
rownames(freq_KNN_sep) <- c("Euro","Aust","Mel","NAm")
write.csv(freq_KNN_sep, "Freq_KNN_sep.csv")

## Calculate Sample sizes
samp_Euro_KNN_sep <- colSums(!is.na(Euro_sim[,c(2:9)]))
samp_Aust_KNN_sep <-  colSums(!is.na(Aust_sim[,c(2:9)]))
samp_Mel_KNN_sep <-  colSums(!is.na(Mel_sim[,c(2:9)]))
samp_NAm_KNN_sep <-  colSums(!is.na(NAm_sim[,c(2:9)]))

samp_KNN_sep <- rbind(samp_Euro_KNN_sep,samp_Aust_KNN_sep,samp_Mel_KNN_sep,samp_NAm_KNN_sep)
colnames(samp_KNN_sep) <- vars
rownames(samp_KNN_sep) <- c("Euro","Aust","Mel","NAm")

write.csv(samp_KNN_sep, "Samp_KNN_sep.csv")

## MMD For KNN_sep Data
MMD_KNN_sep <- read.csv("MMD_KNN_sep.csv")
dend_KNN_sep <- hclust(as.dist(MMD_KNN_sep[,c(2:5)]), method='ward.D2')     
p4 <- ggdendrogram(dend_KNN_sep) + ylim(0,1) + labs(title="KNN Separately")
p4
```

Now, we can grid the plots together to compare the four plots!

```{r Dend Comparison}
grid.arrange(p1,p2,p3,p4,ncol=2)

```





